---
title: "Clustering in R Quick Reference"
author: "Mark Niemann-Ross"
date: "2/6/2021"

output:
  html_document:
    df_print: paged
    toc: yes
  github_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(cluster)

```


This Quick Reference is supplemental to courses on [LinkedIn Learning](https://www.linkedin.com/learning/r-for-data-science-lunchbreak-lessons/).

The latest version of this quick reference is found [here](http://niemannross.com/link/rclusterqref)

A quick reference to [plotting functions in base R is here](http://niemannross.com/link/rplotqref)

A quick reference to [matrix math functions in R is here](http://niemannross.com/link/rmatrixqref)

An index to all R functions covered at LinkedIn Learning is found [here](http://niemannross.com/link/rindex)

[The cluster package](https://cran.r-project.org/web/packages/cluster/index.html) is recommended for advanced operations. The cluster package is based on [Finding Groups in Data](https://onlinelibrary.wiley.com/doi/book/10.1002/9780470316801) 

# Partitioning methods
Divide dataset into k clusters. (k is defined by researcher)

## *pam* : Partitioning Around Medoids

```{r, echo=TRUE}
pam(agriculture, k=2)
```


```{r, echo=TRUE}
plot(pam(agriculture, k=2))
```

## *kmeans* :
## *clara* : Clustering Large Applications
## *fanny* : Fuzzy Analysis Clustering


# Hierarchical methods
Start with cluster of one item, then build larger clusters (or the opposite)

## *agnes* : Agglomerative Nesting

```{r, echo=TRUE}
agnes(agriculture)
```


```{r, echo=TRUE}
plot(agnes(agriculture), which.plots = 2)
```

## *hclust* : 
## *diana* : DIvisive ANAlysis Clustering
## *mona* : MONothetic Analysis Clustering of Binary Variables

# Dissimilarity

## *daisy* : Dissimilarity Matrix Calculation

daisy creates a dissimilarity matrix
```{r, echo=TRUE}
daisy(head(agriculture))
```





