---
title: "300 lessons later: What I've learned about R from teaching R"
author: "Mark Niemann-Ross"
date: "10/13/2020"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

## Slide with Plot

```{r pressure}
plot(pressure)
```

## Snappy Quote

Robert Heinlein said "When one teaches, two learn" - and I'm proof! 

I started learning R in 2014, taught my first R course in 2015. I started with a casual knowledge of data science, statistics, and matrix math. Since then, have taught more than 300 sessions to nearly 60,000 learners. I've covered matrices, NLP, linear models, program flow and other details of the R language. 

 But enough about me - how about you? Let's discuss the objections I've heard about passing on your knowledge to the next generation. 

I'll cover the imposter syndrome, lack of venue, and "I'm a researcher, not an educator." 

In thirty minutes, you'll have your confidence restored and your awesomeness confirmed. 

Mark Niemann-Ross is an author & educator. He teaches for LinkedIn Learning, focused on the R programming language and Raspberry Pi. He is the author of "Stupid Machine" - a science fiction murder mystery solved by a refrigerator.
