---
title: "Plotting in R Quick Reference"
author: "Mark Niemann-Ross"
date: "2/6/2021"

output:
  html_document:
    df_print: paged
    toc: yes
  github_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lattice)
data("ChickWeight")

```

# About
## About This Document

This document provides brief explanations of the plotting functions available in Base R and the lattice package. It does not cover the [tidyverse or ggplot](https://rstudio.com/resources/cheatsheets/) - Rstudio does a great job of that without my help.

## Related Material
This Quick Reference is supplemental to courses on [LinkedIn Learning](https://www.linkedin.com/learning/r-for-data-science-lunchbreak-lessons/). A quick reference to [matrix math functions in R is here](http://niemannross.com/link/rmatrixqref). A quick reference to [clustering functions in R is here](http://niemannross.com/link/rclusterqref). An index to all R functions covered at LinkedIn Learning is found [here](http://niemannross.com/link/rindex). The latest version of this quick reference is found [here](http://niemannross.com/link/rplotqref). The source to this document is found on [github/mnr](https://github.com/mnr/R-for-Data-Science-Lunchbreak-Lessons). This document is available as Free Software under the terms of the [Free Software Foundation’s](http://www.gnu.org/) [GNU General Public License](https://www.r-project.org/COPYING).

## About Mark Niemann-Ross
[Mark](https://niemannross.com/) is an author for LinkedIn Learning and writes Science Fiction.

# plot

[Instructional video on plot()](https://linkedin-learning.pxf.io/plot)

The simplest of all plots

```{r, echo=TRUE}
plot(ChickWeight$Time,ChickWeight$weight)
```


Change the appearance of the plotted points
```{r, echo=TRUE}
plot(ChickWeight$Time,ChickWeight$weight,type = "l")
```

setting the titles and labels

```{r, echo=TRUE}
plot(ChickWeight$Time,ChickWeight$weight,
     type = "l",
     main = "Plot for R Example",
     sub = "We are learning a little at a time",
     xlab="number of days since birth",
     ylab="grams of body weight")
```



add another line to a plot
```{r, echo=TRUE}
plot(ChickWeight$Time,ChickWeight$weight,
     type = "l",
     main = "Plot for R Example",
     sub = "We are learning a little at a time",
     xlab="number of days since birth",
     ylab="grams of body weight")

lines(c(1,20),c(350,50))

```


change the appearance of the graph
```{r, echo=TRUE}

plot(ChickWeight$Time,ChickWeight$weight,
     type = "l",
     main = "Plot for R Example",
     sub = "We are learning a little at a time",
     xlab="number of days since birth",
     ylab="grams of body weight")
lines(c(1,20),c(350,50), type="o", pch=22, lty=2, col="red")

```


You can also use functions as x and y arguments

```{r, echo=TRUE}
some.values <- function(my.something) { 
  return(seq(from=1, to=100,length.out = length(my.something))^2)
  }
plot(mtcars$mpg,some.values(mtcars$mpg))

```


Here's a clever trick - put a plot in a function
```{r, echo=TRUE}

plotThis <- function(lineX,lineY) {plot(mtcars$mpg,
     type = "b",
     main = "Plot for R Example",
     sub = "We are learning a little at a time",
     xlab="mpg is Miles Per Gallon",
     ylab="Look at the pretty bumps")
lines(lineX,lineY)
}
plotThis(c(5,30),c(15,30))

plotThis(c(15,30),c(25,5))

```




# conditional density plot

[Instructional video on cdplot()](https://linkedin-learning.pxf.io/rweekly_cdplot)

Suppose you are doing research: given a certain amount of time, how much should a chick weigh?

```{r echo=TRUE}

# because conditional density plots require a factor
ThreeWeights <- cut(ChickWeight$weight, 3, labels = c(34, 148, 260))

cdplot(ChickWeight$Time, ThreeWeights)

cdplot(ChickWeight$Time, ThreeWeights,
       main = "How much should a chick weigh?",
       ylab = "Probable Weight",
       xlab = "Days"
)

# can you say "overplotting"
cdplot(factor(weight) ~ Time, data = ChickWeight)

# let's fix the overplot with cut()
cdplot(factor(cut(weight, 6, labels = 1:6 * 62)) ~ Time, 
       data = ChickWeight,
       main = "How much should a chick weigh?",
       ylab = "Weight",
       xlab = "Days"
       )

```

# boxplot

[Instructional video on boxplot()](https://linkedin-learning.pxf.io/rweekly_boxplot)

```{r, echo=TRUE}
boxplot(mtcars$mpg)
boxplot(mpg ~ cyl, data = mtcars, col = "lightgray", varwidth = TRUE, 
        main = "mpg vs cylinders",
        ylab = "mpg",xlab = "cylinders")

# Here are the numbers used to create the boxplot
fivenum(mtcars$mpg) 

```


# Histogram

[Instructional video on hist()](https://linkedin-learning.pxf.io/rweekly_histogram)

```{r, echo=TRUE}


hist(ChickWeight$weight)

hist(ChickWeight$weight, density = 30)

hist(ChickWeight$weight, density = 30, breaks = c(0,100,200,max(ChickWeight$weight)))

hist(ChickWeight$weight, breaks = fivenum(ChickWeight$weight))

```

# coplot

[Instructional video on coplot()](https://linkedin-learning.pxf.io/rweekly_coplot)

```{r, echo=TRUE}
coplot(ChickWeight$weight ~ ChickWeight$Time | ChickWeight$Diet)

coplot(weight ~ Time | Diet, data = ChickWeight)

coplot(weight ~ Time | Diet, data = ChickWeight, columns = 4)

coplot(weight ~ Time | Diet, data = ChickWeight, 
       panel = function(x,y,...) { abline(lm(y ~ x)) })

# further explanations:
# http://geog.uoregon.edu/GeogR/topics/coplots.html

```


# barplot

[Instructional video on barplot()](https://linkedin-learning.pxf.io/rweekly_barplot)

```{r, echo=TRUE}

fiveValues <- fivenum(ChickWeight$weight) # just to see what we're working with

barplot(fiveValues) # the simplest of bar plots

barplot(height = fiveValues,
        names.arg = fivenum(ChickWeight$weight),
        horiz = TRUE,
        col = fiveValues,
        main = "Range for Chick Weights"
        )

```

# piechart

[Instructional video on pie()](https://linkedin-learning.pxf.io/rweekly_piechart)

```{r, echo=TRUE}
# simple example of a pie chart
pie(table(ChickWeight$Diet))

# same pie chart with labeled segments
pie(table(ChickWeight$Diet),labels = c("horsebean","linseed","soybean", "sunflower"))

# same pie chart, labeled and with gradients instead of colors
pie(table(ChickWeight$Diet),
    labels = c("horsebean","linseed","soybean", "sunflower"),
    density = 10*1:4,
    angle = 100*1:4)


```

# spineplot

[Instructional video on spineplot()](https://linkedin-learning.pxf.io/rweekly_spineplot)

```{r, echo=TRUE}
# width of bars = frequency of X
# height of bars = frequency of y

# Y must be a factor and is the dependent variable
spineplot(ChickWeight$weight, ChickWeight$Diet) # spineplot(x,y)

# interesting observations
# Height of bars indicates obs per diet. Diet 1 has more obs
# Width of bars indicates obs per weight. More chicks are weighed between 50 and 100

# or...

spineplot(Diet ~ weight, data = ChickWeight) # spineplot(y ~ x)

# bells and whistles

spineplot(Diet ~ weight, 
          data = ChickWeight,
          breaks = fivenum(ChickWeight$weight),
          col = c(5:8),
          xlab = "Chicken Weight",
          ylab = "Chicken Diet")

# The above is actually a spinogram - like a histogram
# example of a true spine plot. Both x and y must be factors
spineplot(Diet ~ factor(weight), 
          data = ChickWeight, 
          col = c(5:8))

spineplot(factor(weight) ~ Diet, 
          data = ChickWeight,
          col = c(1:nlevels(factor(ChickWeight$weight))))

# subset of data
# use the 1st half of data. Would make sense to have a more sophisticated selection
spineplot(Diet ~ factor(weight), 
          data = ChickWeight, 
          subset = c(1:(578/2)),
          drop.unused.levels = TRUE,
          col = c(4:7)
)


```


# dotchart

[Instructional video on dotchart()](https://linkedin-learning.pxf.io/rweeklydotchart)

```{r, echo=TRUE}
# dotchart works with matrix or vector (not dataframe)
# Dotplot for vectors ----
vectorToPlot <- c(1:6)

names(vectorToPlot) <- c(LETTERS[1:6])

dotchart(vectorToPlot, cex = .7)

myGroup <- factor(c("group1","group3","group2","group1","group3","group2"))

dotchart(vectorToPlot, groups = myGroup)

dotchart(vectorToPlot,
         gcolor = "red", groups = myGroup, 
         gdata = c(median(vectorToPlot[myGroup == "group1"]),
                   median(vectorToPlot[myGroup == "group2"]),
                   median(vectorToPlot[myGroup == "group3"])),
         cex = .7,
         main = "Groups of Things", xlab = "Things")

# dotplot for matrix ----

str(WorldPhones) # worldphones is a matrix - not a dataframe
# Major labels (Groups) are matrix columns. Minor labels are matrix rows

dotchart(WorldPhones) # works, but it's messy

dotchart(WorldPhones, gcolor = "Blue", cex = .5,
         gdata = colMeans(WorldPhones), gpch = 15,
         main = "World Phones by Country") 

```


# fourfoldplot

[Instructional video on fourfoldplot()](https://linkedin-learning.pxf.io/rweekly_fourfoldplot)

```{r, echo=TRUE}
# Basic Idea ----
matrixOfData <- matrix(c(2,4,8,32), nrow = 2, ncol = 2, 
                     dimnames = list(c("Chickens","Fish"), c("Big","Small")))

fourfoldplot(matrixOfData)
fourfoldplot(matrixOfData, conf.level = .5, space = .25)

# Compare more than a matrix ----
arrayOfData <- array(seq(from = 1, by = 134, length.out = 12), 
                     dim = c(2,2,3), 
                     dimnames = list(c("Chickens","Fish"), c("Big","Small")))

fourfoldplot(arrayOfData, conf.level = .95, color = c("red","green"),
             std = "all.max",
             main = "Chicken vs Fish",
             mfrow = c(1,3))

```

# matplot

[Instructional video on matplot()](https://linkedin-learning.pxf.io/rweek_matplot)

```{r, echo=TRUE}
# first create some sample matrix
matrix1 <- matrix(c(1:8,1:8), nrow = 8, ncol = 2 )
matrix2 <- matrix(sample(1:16, 16), nrow = 8, ncol = 2)

devAskNewPage(ask = FALSE) # turns off interactive

# then plot. Plots 1st column of 1st matrix against 2nd column of 2nd matrix
matplot(matrix1, matrix2)
matplot(matrix1, matrix2, type = "l")
matplot(matrix1, matrix2, type = "ls") # other types: p, o, b, h, s
matplot(matrix1, matrix2, type = "ls", 
        lty = c(5,6), lwd = c(3,10), lend = c(.1,.3))

#adding a second line
matplot(matrix1, matrix2)
matlines(matrix(c(1:16), nrow = 8, ncol = 2),
         matrix(c(15:0), nrow = 8, ncol = 2))

```


# mosaicplot

[Instructional video on mosaicplot()](https://linkedin-learning.pxf.io/rweekly_mosaicplot)

```{r, echo=TRUE}
# build an array for demonstration
sample1 <- as.integer(runif(10, min = 1, max = 30))
sample2 <- c(10:1)
sample3 <- sample(1:30, 10)

AnimalPlanet <- array(c(sample1, sample2, sample3), 
                 dim = c(2,5,3),
                 dimnames = list(
                   "Planet" = c("Earth", "Mars"),
                   "weight" = c("Featherweight", "Light", "Normal", "Medium", "Heavy" ),
                   "animals" = c("Chickens", "Fish", "Pigs")
                   )
                 )

mosaicplot(AnimalPlanet)


# playin with options
mosaicplot(AnimalPlanet, shade = TRUE)
mosaicplot(AnimalPlanet, shade = fivenum(AnimalPlanet))

mosaicplot(AnimalPlanet, shade = TRUE, sort = c(3,2,1))
mosaicplot(AnimalPlanet, shade = TRUE, off = c(20,15,20))
mosaicplot(AnimalPlanet, shade = TRUE, dir = c("h","v","h"))
mosaicplot(AnimalPlanet, color = TRUE)
mosaicplot(AnimalPlanet, color = c("red","green","blue"))
mosaicplot(AnimalPlanet, cex.axis = .5)
mosaicplot(AnimalPlanet, las = 1) # 0, 1, 2, 3
mosaicplot(AnimalPlanet, type = "f")

# also uses ~ for formula substitution

```


# stemplot

[Instructional video on stem()](https://linkedin-learning.pxf.io/rweekly_stemplot)

```{r, echo=TRUE}

sample1 <- rnorm(100, mean = 10, sd = 3) # creates a normal distribution (bell curve)

sample1 # difficult to parse this data

# doesn't clearly indicate anything
plot(sample1) 

stem(sample1) # demonstrates a normal curve

stem(sample1, scale = 2)

```

# stripchart

[Instructional video on stripchart()](https://linkedin-learning.pxf.io/rweekly_stripchart)

```{r, echo=TRUE}
sample1 <- rnorm(100, mean = 10, sd = 3) # creates a normal distribution (bell curve)

stripchart(sample1)

# labeled multiple strip charts
sample2 <- c(10:1)
sample3 <- sample(1:30, 10)

stripchart(list("Apples" = sample1, "Bananas" = sample2, "Coconuts" = sample3) )

# using the "~" formula
groupFactor <- factor(rep(c("Red", "Blue", "Yellow", "Red"), 25)) # set up a factor

stripchart(sample1 ~ groupFactor)
stripchart(sample1 ~ groupFactor, group.names = c("R", "B", "Y"))

# other
stripchart(sample1, method = "jitter", jitter = 1)
stripchart(sample1, vertical = TRUE)

```


# sunflower

[Instructional video on sunflowerplot()](https://linkedin-learning.pxf.io/rweekly_sunflower)

```{r, echo=TRUE}

xpos <- c(1,2,3,3,1,2,3,1,2,3,4,5,6) # 13 xy points
ypos <- c(1,1,1,1,1,1,1,1,1,1,1,1,1)

# but only 6 points are plotted
plot(xpos, ypos)

# sunflowerplot will show duplicates and the extent they are duplicated
# use xyTable to calculate overlapping points
xyTbl <- xyTable(xpos, ypos)
# then use sunflowerplot to show the overlap at each point
sunflowerplot(xyTbl)

```

# xyplot

[Instructional video on xyplot()](https://linkedin-learning.pxf.io/rweeklatticexyplot)

Part of the lattice package

```{r, echo=TRUE}

# xyplot is an important place to start - all attributes are defined in this help file.

# so easy to set up plots once you understand formulas
xyplot(weight ~ Time, data = ChickWeight)
xyplot(weight*3 ~ Time, data = ChickWeight)
xyplot(weight ~ Time | Diet, data = ChickWeight)

# what if you want to add two variables instead of combine graphs?
ChickWeight$anotherVar <- 1:nrow(ChickWeight) # add a variable

# not what we want
xyplot(weight ~ Time + anotherVar, type = "l", data = ChickWeight)
# use I() (inhibit interpretation) to get desired result
xyplot(weight ~ I(Time + anotherVar), type = "l", data = ChickWeight)


# lots of options for creating keys/legends
xyplot(weight ~ Time | Diet, 
       data = ChickWeight,
       type = "a", # lines = average
       groups = Diet, # necessary for auto.key,
       auto.key = list(columns = 2, 
                       points = FALSE,
                       rectangles = TRUE,
                       space = "bottom"))

# or just use autokey
xyplot( weight ~ Time, 
        groups = Diet,
        data = ChickWeight, 
        type = "l",
        auto.key = TRUE)

```



# cloud and wireframe

[Instructional video on cloud()](https://linkedin-learning.pxf.io/rweeklycloudplot)

Part of the lattice package

```{r, echo=TRUE}

# clouds and wireframes "look" the same...but

# I need a third numeric variable
ChickWeight$ascend <- 1:nrow(ChickWeight)
names(ChickWeight) # review names of columns

cloud( weight ~ Time * ascend, data = ChickWeight)

# labeling axis
cloud( weight ~ Time * Diet, 
       data = ChickWeight,
       scales = list(arrows = FALSE))

# wireframe focused on comparing two variables against a third
wireframe( weight ~ Time * Diet, data = ChickWeight)

# also can be done with a matrix

mymatrix <- matrix(1:10000, nrow = 100)
wireframe(mymatrix)

# ooh - fancy!
wireframe(mymatrix,
          shade = TRUE,
          light.source = c(10,0,10),
          drape = TRUE
)


# more parameters in xyplot and wireframe documentation

```

# level and contour

[Instructional video on levelplot() and contourplot()](https://linkedin-learning.pxf.io/rweeklycontourplot)

Part of the lattice package

```{r, echo=TRUE}

names(ChickWeight) # review names of columns

# using exact same setup as cloud and wireframe
levelplot( weight ~ Time * Diet, data = ChickWeight)
levelplot( weight ~ Time * Diet, data = ChickWeight, contour=TRUE) # adds lines
levelplot( weight ~ Time * Diet, data = ChickWeight, cuts=7) # default cuts = 15

contourplot( weight ~ Time * Diet, data = ChickWeight)
contourplot( weight ~ Time * Diet, data = ChickWeight, region=TRUE) # adds levels
contourplot( weight ~ Time * Diet, data = ChickWeight, labels = FALSE) # removes labels from lines
contourplot( weight ~ Time * Diet, data = ChickWeight, cuts=30) 


```

# barchart

[Instructional video on barchart()](https://linkedin-learning.pxf.io/rweeklybarchart)

Part of the lattice package

```{r, echo=TRUE}

# barchart using standard formula setup

barchart( weight ~ Time + Diet, data = ChickWeight)

barchart( Diet ~ weight, data = ChickWeight) # these plots make more sense with a factor 

# With barchart options
barchart( weight ~ Time + Diet, data = ChickWeight)
barchart( weight ~ Time + Diet, data = ChickWeight, box.ratio = 10)
barchart( weight ~ Time + Diet, data = ChickWeight, horizontal = TRUE) # note the axis stay the same


```


# bwplot

[Instructional video on bwplot()](https://linkedin-learning.pxf.io/rweeklybarchart)

```{r, echo=TRUE}
bwplot( Diet ~ Time, data = ChickWeight)


# with bwplot options
bwplot( Diet ~ Time, data = ChickWeight, box.ratio = 10)
bwplot( Diet ~ Time, data = ChickWeight,
        notch = TRUE, varwidth = TRUE
)

# changing axis label
levels(ChickWeight$Diet) <- c("Apples","Bananas","Steak","Potatoes")
bwplot( Diet ~ Time, data = ChickWeight)

```


# splom

[Instructional video on splom()](https://linkedin-learning.pxf.io/rweeklysplom)

Part of the lattice package

SPLOM == Scatter PLot Matrices

```{r, echo=TRUE}
splom(ChickWeight)

splom(ChickWeight[1:2]) # plots weight & time

splom(~ChickWeight[1:2], 
      groups = Diet, 
      data = ChickWeight)

# smooths out the scatter plot
splom(~ChickWeight[1:2], 
      groups = Diet, 
      data = ChickWeight,
      panel = panel.smoothScatter )


# colorful - but is it useful?
parallelplot(~ChickWeight[1:2], 
      groups = Diet, 
      data = ChickWeight) 

```


# panels

[Instructional video on lattice panels](https://linkedin-learning.pxf.io/rweekpanels)

Part of the lattice package

```{r, echo=TRUE}
# changing graph to smoothscatter plot
splom(~ChickWeight[1:2], 
      groups = Diet, 
      data = ChickWeight,
      panel = panel.smoothScatter)

# other panel options
splom(~ChickWeight[1:2], 
      data = ChickWeight,
      panel = panel.loess)

splom(~ChickWeight[1:2], 
      data = ChickWeight,
      panel = panel.spline)

# change bwplot to a violin plot
bwplot( Diet ~ Time, data = ChickWeight, 
        panel = function(..., box.ratio) {
          panel.violin(..., col = "transparent",
                       varwidth = TRUE, box.ratio = box.ratio)
          panel.bwplot(..., fill = NULL, box.ratio = .1)
          }
          )

# combine panels

someColors <- colorRampPalette(c("yellow","blue"))

# create a function to be used with panels
# look up ?panel for more options
myPanelFunc <- function(x,y,..){
  panel.smoothScatter(x,y, lwd = 1, colramp = someColors)
  panel.loess(x,y, lwd = 2, col = "red")
}

#...then set panel = to the panel plotting function
splom(~ChickWeight[1:2], 
      groups = Diet, 
      data = ChickWeight,
      panel = myPanelFunc
)

# list of plot types and possible panels

# barchart: panel.barchart
# bwplot: panel.bwplot, panel.violin
# cloud : panel.cloud, panel.3dscatter, panel.3dwire, panel.wireframe
# contourplot: panel.contourplot, panel.levelplot, panel.levelplot.raster
# densityplot: panel.densityplot
# dotplot: panel.dotplot
# generics: panel.abline, panel.arrows, panel.average, panel.axis, panel.curve, 
#           panel.fill, panel.functions, panel.grid, panel.linejoin, panel.lines
#           panel.lmline, panel.loess, panel.mathdensity, panel.points, panel.polygon,
#           panel.refline, panel.rug, panel.smoothscatter, panel.violin
# histogram: panel.histogram
# levelplot: panel.contourplot, panel.levelplot, panel.levelplot.raster
# parallel: panel.parallel
# qqmath: panel.qqmath, panel.qqmathline
# stripplot: panel.stripplot
# xyplot: panel.xyplot, panel.qq, panel.splom


```

# strip plot

[Instructional video on stripplot()](https://linkedin-learning.pxf.io/rweekstripplot)

Part of the lattice package

```{r, echo=TRUE}

# simple plot. Pay attention to formula factor~numeric
stripplot( Diet ~ weight , data = ChickWeight)

# with panel.stripplot options
# note these can be passed without using panel.stripplot()

# jitter
stripplot( Diet ~ weight , data = ChickWeight,
           jitter.data = TRUE
           )

stripplot( Diet ~ weight , data = ChickWeight,
           jitter.data = TRUE,
           factor = 1
)

stripplot( Diet ~ weight , data = ChickWeight,
           jitter.data = TRUE,
           factor = .5
)

stripplot( Diet ~ weight , data = ChickWeight,
           jitter.data = TRUE,
           amount = .2 # similar to factor
)


# groups
stripplot( Diet ~ weight , data = ChickWeight,
           jitter.data = TRUE,
           factor = 1,
           groups = Diet
)

# not what we want
stripplot( Diet ~ weight , data = ChickWeight,
           horizontal = FALSE
)

# Need to also change the formula
stripplot( weight ~ Diet , data = ChickWeight,
           horizontal = FALSE
)

```

